---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First, download and put the csv file in our working directory. Then read the file.
```{r, echo = TRUE}
d <- read.csv("activity.csv", header = TRUE)
attach(d)
```
The data is loaded. I will directly use this data frame without any preprocessing.


## What is mean total number of steps taken per day?
I use the aggregate function to find the total number of steps taken per day and make a data frame "day_total" which contains the total number of steps per day.

```{r, echo = TRUE}
day_total <- aggregate(list(total.steps = steps), 
                       list(date = date), FUN=sum)
head(day_total)
```
Now we will make a histogram of the total number of steps taken each day.
```{r , echo= TRUE}
total.steps <- day_total$total.steps
hist(total.steps, breaks = 10)
```

Now I will print the mean and median of the total number of steps taken per day.
```{r, echo=TRUE}
summary(total.steps)["Median"]
summary(total.steps)["Mean"]
```
## What is the average daily activity pattern?
Now I will find the daily activity pattern. I will use the aggregate function. 
```{r, echo = TRUE}
pattern <- aggregate(list(daily.avg = steps), 
                     list(interval = interval), 
                     FUN = mean, na.rm = TRUE)
plot(pattern$interval, pattern$daily.avg, 
     type="l", xlab = "interval", ylab = "daily.avg")
```

I will also find the interval when the daily average is the maximum.
```{r, echo=TRUE}
pattern[which.max(pattern$daily.avg),1]
```
## Imputing missing values
Now I will find the total number of missing values in the dataset.
```{r, echo=TRUE}
missing <- sum(is.na(steps))
sprintf("Total number of missing values: %s", 
        missing)
```
Now I will imput values for the na s. I will substitute all the na's with the daily average for that 5-minute interval.
```{r, echo = TRUE}
new_df <- data.frame(date= date,
                     interval = interval)

new_steps <- steps
for (i in 1:length(steps))
    { if (is.na(steps[i]))
    {new_steps[i] <- 
        pattern$daily.avg[interval[i]==pattern$interval]}
}
new_df["new_steps"] <- new_steps
head(new_df)
```

With this new dataframe I will plot the histogram and time series again.
```{r, echo = TRUE}
par(mfrow = c(1,2))
newday_total <- aggregate(list(total.steps = new_steps), 
                          list(date = date), FUN=sum, na.rm=TRUE)
head(newday_total)
newtotal.steps <- newday_total$total.steps
hist(newtotal.steps, 
     breaks = 10, main = "Imputed values")
hist(total.steps, breaks = 10, 
     main = "Original values")
summary(newtotal.steps)["Median"]
summary(newtotal.steps)["Mean"]
```
We see that the median remains same but the mean increases. We can easily see that there are a lot of values in the left side of the histogram, because we ignored the missing values. But after imputing the missing values the histogram now looks a bell shaped histogram.
## Are there differences in activity patterns between weekdays and weekends?
```{r, echo = TRUE, message=FALSE}
library(chron)
library(dplyr)
library(lattice)
new_df$weekdays <- ifelse(is.weekend(new_df$date),
                          "Weekend","Weekday")
head(new_df)
daily <- group_by(new_df, interval, weekdays)
daily.data <-  summarise(daily, daily.avg = mean(new_steps))
xyplot(daily.data$daily.avg~ daily.data$interval|daily.data$weekdays,
       type = "l", layout = c(1,2),
       xlab= "interval", ylab = "daily average", 
       main = "weekdays vs weekend")
```

